Spectral Experts for Estimating Mixtures of Linear Regressions

### 2013-06-04 doc/icml-presentation/presentation.tex
Marked out the flow of the presentation

\begin{frame}
  \frametitle{Discriminative Latent Variable Models}

  \begin{itemize}
    \item {\bf Discriminative Latent Variable Models are useful, but hard to learn.}
    \item Discriminative LVMs combine the predictive accuracy with compact expressiveness.
    \item They have been used in several applications, e.g.\ object recognition, syntactic parsing, machine translation, etc.
    \item Learning parameters for these models is hard because they have non-convex likelihood functions.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Goal: Consistent Parameter Estimation.}

  \begin{itemize}
    \item {\bf We would like to develop efficient consistent estimators for discriminative LVMS.}
    \item Past approaches rely on local optimization, which are susceptible to local optima.
    \item And they don't just go away with more data.
    \item Recently, there has been work on consistent parameter
      estimation for generative LVMS\@. We'd like to extend their work to
      the discriminative case.
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
    \item {\bf First step, consider a simple discriminative model, mixture of linear regressions.}
    \item The mixture of linear regressions model defines a conditional distribution over a response $y \in \Re$ given covariates $x \in \Re^d$.
     \item The generative procedure is as follows,
       \begin{enumerate}
        \item Draw a mixture component $h \in [k] \sim Mult(\pi)$, where $\pi = [\pi_1 | \dots | \pi_k]$ defines the mixture proportions.
        \item Draw the noise $\epsilon \sim \mathcal{E}$, where $\mathcal{E}$ is the noise distribution.
        \item Set $y = \beta_h^T x + \epsilon$, where $\{\beta_h\}_{h=1}^{k}$ are the conditional means of the regression coefficients.
       \end{enumerate}
    \item The parameters that we would like to learn from this model are $\pi$ and $B = [ \beta_1 | \cdots | \beta_k ]$.
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
    \item {\bf We approach the problem by using the method of moments on three conditional views. }
    \item Consider a moment map $\mathcal{M}$ that maps the parameters $\theta$ to the moments $m$. For a Gaussian, we have that $\mathcal{M} = (\mu, \sigma^2)$.
    \item In general, we will compute the inverse of the moment map to learn the parameters from the sample estimates.
    \item By the central limit theorem, our sample estimates of the moments converge at a $1/\sqrt{n}$ rate, so we expect that our parameters will also converge at this rate.
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
    \item {\bf We can recover the means of a generative LVM by exploiting their symmetric structure. }
    \item 
    \begin{align*}
    M_2 &= \sum_{h=1}^k \pi_h \beta_h\tp{2} & M_3 &= \sum_{h=1}^k \pi_h \beta_h\tp{3}.
  \end{align*}
    \item In general, $M_2$ is insufficient to identify the model, because it is invariant to rotations of $B$. 
    \item If the $\beta_h$ were orthogonal, then the eigenvectors and eigenvalues of $M_3$ would give us the answer. 
    \item In general, we need both $M_2$ and $M_3$. We can use the whitening transformation for $M_2$ to whiten $M_3$ such that it has a orthogonal decomposition.
\item The robust tensor power method by \cite{AnandkumarHsuGe2012} find stable eigenvectors.
  \item In the discriminative case, we do not observe $M_2$ directly from the data, but rather, as we'll see, linear projections of them.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Recovering the Moments.}
  \begin{itemize}
    \item {\bf The powers of the response variables $y^p$ are linear measurements of the moments of the parameters $M_p$!}
    \item As noted earlier, the first problem we run into is that we can't observe the moments of the parameters $B$ and $\pi$ directly!
\item However, observe that 
\begin{align}
  y &= \innerp{\beta_h}{x} + \epsilon \\
    &= \innerp{M_1}{x} + \underbrace{{\beta_h - M_1}{x} + \epsilon }_{\textrm{noise}},
\end{align}
  where $M_1 = \sum_{h=1}^k \pi_h \beta_h$, the mean regression coefficient. 
  \item We note that while the noise term is dependent on $x$, it has a zero-mean. Thus, we could potentially recover $M_1$ through regression.
\item Similarly, we can cast the remaining two terms as regression problems.

\item An additional fact that we can exploit is that both $M_2$ and $M_3$ are low rank, so we can use low-rank regression to recover estimates $\hat M_2$ and $\hat M_3$ efficiently from data.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Rates for Recovery}
  \begin{itemize}
    \item {\bf The algorithm is consistent, and requires $O(?)$ samples.}
\item The rate of convergence for the spectral experts algorithm to the
true parameters breaks into two parts; the rates for learning the
moments, which feeds into the rates for learning the parameters.
\item For low rank regression, we have the following bound on recovery by (Tomioka2011);
  $$ \| \hat M_p - M_p \|_F \le \frac{32 \lambda^{(p)}_n \sqrt{k}}{\kappa(\opX_p)}, $$
where $\kappa(\opX_p)$ is the (restricted) strong convexity constant, and $\lambda^{(p)} > \|\opX^*_p(\eta)\|$. 
\item Because we assume our noise is bounded, it is easy to show that the error concentrates. 
\item In the tensor recovery case, we will need to whiten $M_3$ before
applying the tensor decomposition and unwhiten it afterwards; this
modifies the error bounds slightly.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Experimental Results}
  \begin{itemize}
    \item {\bf Spectral Experts provides a good initialization to EM.}
      \item We simulated the performance of spectral experts and compared it to EM. 
\item In a non-trivial number of cases, EM did not converge to the right parameters and got stuck in local optima.
\item The parameters estimated by the spectral method were not very good even with $O(10^5)$ samples.
\item However, EM initialized with these parameters did extremely well. 
\item This finding that spectral methods should be a good initialization for EM is not surprising.
  \begin{itemize}
    \item The biggest sell for spectral methods is that they give a global guarantee on where the parameters are.
    \item The parameters might not be at the global optima, but will hopefully lie in the potential well around it.
    \item EM will then converge to the global optima.
  \end{itemize}
  \end{itemize}
\end{frame}

>>>>

\begin{frame}
  \frametitle{Generative vs. Discriminative Latent Variable Models}

  \begin{centering}
  \begin{tikzpicture}
    % x, y
    \node at (0,0) {Generative Models};
    \node[style=box] at (0,-1.5) {\obj{%
      \vspace{4em}
      \hfill CRFs
      }};
    \node[style=box] at (0,-4) {\obj{%
      \vspace{4em}
      \hfill PCFGs
      }};

    \node<2-> at (6,0) {Discriminative Models};
    \node<2->[style=box] at (6,-1.5) {\obj{%
      \vspace{4em}
      \hfill Liang et.\ al (2006)
      }};
    \node<2->[style=box] at (6,-4) {\obj{%
      \vspace{4em}
      \hfill Wang \& Mori (2009)
      }};

    \node<2-> at (7,-5.5) {\obj{%
      Easier to include features, \\
      tend to be more accuracy.
      }};
  \end{tikzpicture}
  \end{centering}

    \note[item]{Discriminative Latent Variable Models are useful, but hard to learn.}
    \note[item]{Split models into generative and discriminative models.}
    \note[item]{Discriminative LVMs combine the predictive accuracy with compact expressiveness.}
    \note[item]{They have been used in several applications, e.g.\ object recognition, syntactic parsing, machine translation, etc.}
    \note[item]{Learning parameters for these models is hard because they have non-convex likelihood functions.}
\end{frame}

\begin{frame}
  \frametitle{Parameter Estimation is Hard}

  \begin{centering}
  \begin{tikzpicture}
    % x, y
    \node[style=box] at (0,-1.5) {\objw{36em}{%
      \vspace{10em}
      {\em log-likelihood picture}
      }};
  \end{tikzpicture}
  \end{centering}

  \begin{itemize}
    \item<1-> Log-likelihood function is non-convex.
    \item<2-> This is a problem that doesn't go away even with infinite data.
    \item<3-> Can we build {\bf consistent estimators} that get better with more data?
  \end{itemize}

  \note[item]{We would like to develop efficient consistent estimators for discriminative LVMS.}
  \note[item]{Past approaches rely on local optimization, which are susceptible to local optima.}
  \note[item]{And they don't just go away with more data.}
  \note[item]{Recently, there has been work on consistent parameter
      estimation for generative LVMS\@. We'd like to extend their work to
      the discriminative case.}
\end{frame}

\begin{frame}
  \frametitle{Spectral Methods for Generative Models.}
  \begin{centering}
  \begin{tikzpicture}
    % x, y
    \node at (0,0) {Generative Models};
    \node[style=box] at (0,-1.5) {\obj{%
      \vspace{4em}
      \hfill CRFs
      }};
    \node[style=box] at (0,-4) {\obj{%
      \vspace{4em}
      \hfill PCFGs
      }};

    %\node<2-> at (6,0) {Discriminative Models};
    \node<2->[style=box] at (6,-1.5) {\obj{%
      Solve the general multi-view case using the tensor factorization method.\\
      \hfill Anandkumar et.\ al (2012) 
      }};
    \node<3->[style=box] at (6,-4) {\obj{%
      Reduce the problem to multi-view learning via tree-unmixing.\\
      \hfill Hsu et.\ al (2012) 
      }};
  \end{tikzpicture}
  \end{centering}

   \todo{Talk about observable operator stuff.}
\end{frame}

@@

  \begin{centering}
  \begin{tikzpicture}
    % x, y
    \node<1-3> at (0,0){Generative Process};
    \node<1-3>[style=box] at (0,-2.5) {\obj{%
      Graphical model \vspace{3em}\\
      Given parameters $\pi$ and $B = [ \beta_1 | \cdots | \beta_k ]$,
      \begin{itemize}
        \item<1-> $h \in [k] \sim Mult(\pi)$.
        \item<2-> $\epsilon \sim \mathcal{E}$.
        \item<3-> $y = \beta_h^T x + \epsilon$.
      \end{itemize}
      }};
    \node<4>[style=box] at (0,0) {\obj{%
    {\bf Given:} $\mathcal{D} = (x_i, y_i)_{i=1}^{n}$,\\
    can we {\bf estimate parameters} $\pi$ and $B = [ \beta_1 | \cdots | \beta_k ]$?
      }};

    %\node<2-> at (6,0) {Discriminative Models};
    \node[style=box] at (6,-1.5) {\obj{%
      \vspace{12em}
      Diagram
      }};
  \end{tikzpicture}
  \end{centering}

  \note[item]{First step, consider a simple discriminative model, mixture of linear regressions.}
  \note[item]{The mixture of linear regressions model defines a conditional distribution over a response $y \in \Re$ given covariates $x \in \Re^d$.}
     \note[item]{The generative procedure is as follows,
       \begin{enumerate}
        \item Draw a mixture component $h \in [k] \sim Mult(\pi)$, where $\pi = [\pi_1 | \dots | \pi_k]$ defines the mixture proportions.
        \item Draw the noise $\epsilon \sim \mathcal{E}$, where $\mathcal{E}$ is the noise distribution.
        \item Set $y = \beta_h^T x + \epsilon$, where $\{\beta_h\}_{h=1}^{k}$ are the conditional means of the regression coefficients.
      \end{enumerate}}
   \note[item]{The parameters that we would like to learn from this model are $\pi$ and $B = [ \beta_1 | \cdots | \beta_k ]$.}
\end{frame}

@@

  \begin{centering}
  \begin{tikzpicture}
    % x, y
    \node[style=box] at (0,0) {\objw{36em}{%
      \vspace{10em}
      {\em moment-map picture}
      }};
  \end{tikzpicture}
  \end{centering}
  \begin{itemize}
    \item<1-> Moments are expected to converge at a $1/\sqrt{n}$ rate, by CLT.
    \item<2-> A continuous mapping $\mathcal{M}^{-1}$ will also converge at a $1/\sqrt{n}$ rate.
  \end{itemize}

    \note[item]{We approach the problem by using the method of moments on three conditional views. }
    \note[item]{Consider a moment map $\mathcal{M}$ that maps the parameters $\theta$ to the moments $m$. For a Gaussian, we have that $\mathcal{M} = (\mu, \sigma^2)$.}
    \note[item]{In general, we will compute the inverse of the moment map to learn the parameters from the sample estimates.}
    \note[item]{By the central limit theorem, our sample estimates of the moments converge at a $1/\sqrt{n}$ rate, so we expect that our parameters will also converge at this rate.}
\end{frame}

@@

  \begin{centering}
  \begin{tikzpicture}
    % x, y
    \node[style=box] at (0,0) {\objw{36em}{%
      \vspace{10em}
      {\em Work out the moments in the generative case, with indicative diagrams for matrices, tensors, etc.}
      }};
  \end{tikzpicture}
  \end{centering}

    \note[item]{\bf We can recover the means of a generative LVM by exploiting their symmetric structure.}
    \note[item]{\begin{align*} M_2 &= \sum_{h=1}^k \pi_h \beta_h\tp{2} & M_3 &= \sum_{h=1}^k \pi_h \beta_h\tp{3}.  \end{align*}}
      \note[item]{In general, $M_2$ is insufficient to identify the model, because it is invariant to rotations of $B$. }
      \note[item]{If the $\beta_h$ were orthogonal, then the eigenvectors and eigenvalues of $M_3$ would give us the answer. }
      \note[item]{In general, we need both $M_2$ and $M_3$. We can use the whitening transformation for $M_2$ to whiten $M_3$ such that it has a orthogonal decomposition.}
      \note[item]{The robust tensor power method by \cite{AnandkumarHsuGe2012} find stable eigenvectors.}
\end{frame}

\begin{frame}
  \frametitle{Method of Moments for the Mixture of Linear Regressions.}

  \begin{centering}
  \begin{tikzpicture}
    % x, y
    \node[style=box] at (0,0) {\objw{36em}{%
      \vspace{10em}
      {\em Try to work out the moments in our case.}
      }};
  \end{tikzpicture}
  \end{centering}
      \note[item]{In the discriminative case, we do not observe $M_2$ directly from the data, but rather, as we'll see, linear measurements of them.}
      \note[item]{Try to work out the moments for this case and see how they {\tt \#fail} epically.}
\end{frame}

\begin{frame}
  \frametitle{Method of Moments for the Mixture of Linear Regressions.}

  \begin{centering}
  \begin{tikzpicture}
    % x, y
    \node[style=box] at (0,0) {\objw{18em}{%
      \vspace{10em}
      {\em Different tactic: show that $y^p$ are linear measurements of the moments of the parameters $M_p$! }
      }};

    \node[style=box] at (6,0) {\objw{8em}{%
      {$y^p$ are linear measurements of the moments of the parameters $M_p$!}
      }};
  \end{tikzpicture}
  \end{centering}

    \note[item]{The powers of the response variables $y^p$ are linear measurements of the moments of the parameters $M_p$!}
    \note[item]{As noted earlier, the first problem we run into is that we can't observe the moments of the parameters $B$ and $\pi$ directly!}
    \note[item]{However, observe that 
    \begin{align}
      y &= \innerp{\beta_h}{x} + \epsilon \\
        &= \innerp{M_1}{x} + \underbrace{{\beta_h - M_1}{x} + \epsilon }_{\textrm{noise}},
    \end{align}
    where $M_1 = \sum_{h=1}^k \pi_h \beta_h$, the mean regression coefficient.}
    \note[item]{We note that while the noise term is dependent on $x$, it has a zero-mean. Thus, we could potentially recover $M_1$ through regression.}
    \note[item]{Similarly, we can cast the remaining two terms as regression problems.}
    \note[item]{An additional fact that we can exploit is that both $M_2$ and $M_3$ are low rank, so we can use low-rank regression to recover estimates $\hat M_2$ and $\hat M_3$ efficiently from data.}
\end{frame}

\begin{frame}
  \frametitle{Efficient Consistent Recovery}

  \begin{centering}
  \begin{tikzpicture}
    % x, y
    \node[style=box] at (0,0) {\objw{12em}{%
      \vspace{10em}
      Regression step
      }};

    \node[style=box] at (6,0) {\objw{12em}{%
      The tensor factorization method.
      }};
  \end{tikzpicture}
  \end{centering}

  \note[item]{The algorithm is consistent, and requires $O(?)$ samples.}
  \note[item]{The rate of convergence for the spectral experts algorithm to the true parameters breaks into two parts; the rates for learning the moments, which feeds into the rates for learning the parameters.}
  \note[item]{For low rank regression, we have the following bound on recovery by (Tomioka2011); $$ \| \hat M_p - M_p \|_F \le \frac{32 \lambda^{(p)}_n \sqrt{k}}{\kappa(\opX_p)}, $$ where $\kappa(\opX_p)$ is the (restricted) strong convexity constant, and $\lambda^{(p)} > \|\opX^*_p(\eta)\|$.}
  \note[item]{Because we assume our noise is bounded, it is easy to show that the error concentrates.}
  \note[item]{In the tensor recovery case, we will need to whiten $M_3$ before applying the tensor decomposition and unwhiten it afterwards; this modifies the error bounds slightly.}
\end{frame}

\begin{frame}
  \frametitle{Experimental Insights}
  \begin{itemize}
    \item {\bf Spectral Experts provides a good initialization to EM.}
      \item We simulated the performance of spectral experts and compared it to EM. 
\item In a non-trivial number of cases, EM did not converge to the right parameters and got stuck in local optima.
\item The parameters estimated by the spectral method were not very good even with $O(10^5)$ samples.
\item However, EM initialized with these parameters did extremely well. 
\item This finding that spectral methods should be a good initialization for EM is not surprising.
  \begin{itemize}
    \item The biggest sell for spectral methods is that they give a global guarantee on where the parameters are.
    \item The parameters might not be at the global optima, but will hopefully lie in the potential well around it.
    \item EM will then converge to the global optima.
  \end{itemize}
  \end{itemize}
\end{frame}
